---
title: "Project1 D612"
author: "Jose Fuentes"
date: "2025-06-07"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Project 1 DATA 612

Briefly describe the recommender system that you’re going to build out from a business
perspective, e.g. “This system recommends data science books to readers.”
• Find a dataset, or build out your own toy dataset. As a minimum requirement for complexity,
please include numeric ratings for at least five users, across at least five items, with some missing
data.
• Load your data into (for example) an R or pandas dataframe, a Python dictionary or list of lists, (or
another data structure of your choosing). From there, create a user-item matrix.
• If you choose to work with a large dataset, you’re encouraged to also create a small, relatively
dense “user-item” matrix as a subset so that you can hand-verify your calculations.
• Break your ratings into separate training and test datasets.
• Using your training data, calculate the raw average (mean) rating for every user-item combination.
• Calculate the RMSE for raw average for both your training data and your test data.
• Using your training data, calculate the bias for each user and each item.
• From the raw average, and the appropriate user and item biases, calculate the baseline predictors
for every user-item combination.
• Calculate the RMSE for the baseline predictors for both your training data and your test data.
• Summarize your results.

For this project MovieLens dataset is going to be used and the link is shown as follow, extract a smaller dense matrix from it.

```{r}
# Loading packages
library(tidyverse)
library(knitr)
library(kableExtra)

# Downloading and using  MOVIELENS data
download.file("https://files.grouplens.org/datasets/movielens/ml-100k.zip", 
              "ml-100k.zip")
unzip("ml-100k.zip", files = "ml-100k/u.data")
ratings <- read.table("ml-100k/u.data", 
                      sep = "\t", 
                      col.names = c("user", "item", "rating", "timestamp"))

# Select dense 10x10 matrix
top_users <- ratings %>%
  count(user) %>%
  top_n(10, n) %>%
  pull(user)

top_items <- ratings %>%
  filter(user %in% top_users) %>%
  count(item) %>%
  top_n(10, n) %>%
  pull(item)

dense_ratings <- ratings %>%
  filter(user %in% top_users, item %in% top_items) %>%
  select(user, item, rating)

# train/test split
set.seed(42)
dense_ratings <- dense_ratings %>%
  mutate(split = ifelse(runif(n()) < 0.8, "train", "test"))

train <- dense_ratings %>% filter(split == "train")
test  <- dense_ratings %>% filter(split == "test")

#Global mean function
global_mean <- mean(train$rating)

# User and Item biases
user_bias <- train %>%
  group_by(user) %>%
  summarise(user_bias = mean(rating - global_mean))

item_bias <- train %>%
  group_by(item) %>%
  summarise(item_bias = mean(rating - global_mean))

# Creating Baseline matrix using training data
user_ids <- sort(unique(train$user))
item_ids <- sort(unique(train$item))

bias_grid <- expand.grid(user = user_ids, item = item_ids) %>%
  left_join(user_bias, by = "user") %>%
  left_join(item_bias, by = "item") %>%
  mutate(
    user_bias = replace_na(user_bias, 0),
    item_bias = replace_na(item_bias, 0),
    baseline = global_mean + user_bias + item_bias,
    user = paste0("User_", user)  # ensuring user names are unique
  )

baseline_matrix <- bias_grid %>%
  # Specify that 'user' is the unique identifier, and if duplicates exist, use the first value.
  pivot_wider(
    id_cols = user, 
    names_from = item, 
    values_from = baseline, 
    values_fn = first
  ) %>%
  column_to_rownames("user") %>%
  as.matrix()

# === CREATE TEST MATRIX (FROM TEST DATA) ===
test_matrix <- test %>%
  mutate(user = paste0("User_", user)) %>%
  pivot_wider(
    id_cols = user, 
    names_from = item, 
    values_from = rating, 
    values_fn = mean
  ) %>%
  column_to_rownames("user") %>%
  as.matrix()

#  Aling matrices using common user and items
common_users <- intersect(rownames(baseline_matrix), rownames(test_matrix))
common_items <- intersect(colnames(baseline_matrix), colnames(test_matrix))

baseline_aligned <- baseline_matrix[common_users, common_items, drop = FALSE]
test_aligned     <- test_matrix[common_users, common_items, drop = FALSE]

# === GLOBAL AVERAGE MATRIX FOR REFERENCE ===
raw_matrix <- matrix(global_mean,
                     nrow  = nrow(baseline_aligned),
                     ncol  = ncol(baseline_aligned),
                     dimnames = dimnames(baseline_aligned))

# RMSE Function
rmse <- function(x, y) {
  sqrt(mean((x - y)^2, na.rm = TRUE))
}

# Calculating RMSE using Aligned matrices
rmse_test_raw  <- rmse(test_aligned, raw_matrix)
rmse_test_base <- rmse(test_aligned, baseline_aligned)

# Sumarry table
summary_rmse <- tibble(
  Method   = c("Raw Average", "Baseline Predictor"),
  Test_RMSE = c(round(rmse_test_raw, 3), round(rmse_test_base, 3))
)

kable(summary_rmse, caption = "Summary of RMSEs (Test Set)") %>%
  kable_styling(bootstrap_options = c("striped", "bordered", "hover", "condensed")) %>%
  row_spec(0, bold = TRUE, color = "white", background = "lightblue")

```


## Including Plots

You can also embed plots, for example:

```{r}
# ------------------------------
# Example: Using Training Data to Compare Predicted and Actual Ratings
# and to generate recommendations for a chosen user.
# ------------------------------

# Choose one user from the training set (for example, user ID 13)
chosen_id <- 13
chosen_user <- paste0("User_", chosen_id)  # This is how users are named in baseline_matrix

# Extract predicted (baseline) ratings for the chosen user from baseline_matrix
predicted_ratings <- baseline_matrix[chosen_user, ]
predicted_df <- tibble(
  item = as.integer(names(predicted_ratings)),
  predicted_rating = as.numeric(predicted_ratings)
)

# Get the actual ratings for the chosen user from the training dataset
actual_df <- train %>%
  filter(user == chosen_id) %>%
  select(item, rating)

# Compare predicted versus actual ratings by joining the data
comparison <- left_join(predicted_df, actual_df, by = "item")
cat("Comparison of Predicted Ratings vs. Actual Ratings for", chosen_user, ":\n")
print(comparison)

# ----- Generating Recommendations -----
# Identify items the user has not rated in the training dataset.
# These items are potential recommendations if their predicted
# rating is high.
recommendations <- predicted_df %>%
  anti_join(actual_df, by = "item") %>%
  arrange(desc(predicted_rating))

cat("\nRecommended Items for", chosen_user, "based on Baseline Predictor (unrated in training):\n")
print(recommendations)

```

Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.
